Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016

### 2020-04-05 anzjs_elections.Rmd
Final tense changes

The spatial autoregressive coefficient $\rho$ was positive and significant in the 2001, 2004 and 2016 elections (Figure \@ref(fig:rhovis)). In these three elections, there is evidence to suggest that neighbours shared some influential characteristics outside the explanatory variables, which affected the two-party preferred vote. Conversely, in the other three elections, the spatial effect is weaker and insignificant (although still positive).

```{r rhovis, fig.cap="Estimates of the spatial autoregressive parameter for each of the six elections, reported with their individual 95\\% confidence intervals. In 2001, 2004 and 2016 there is a significant spatial component.", out.height='25%', out.extra = ''}
rho_df <- bind_rows(
  glsmod16$rho_df %>% mutate(year = "2016"),
  glsmod13$rho_df %>% mutate(year = "2013"),
  glsmod10$rho_df %>% mutate(year = "2010"),
  glsmod07$rho_df %>% mutate(year = "2007"),
  glsmod04$rho_df %>% mutate(year = "2004"),
  glsmod01$rho_df %>% mutate(year = "2001")
) %>%
  mutate(
    upper95 = estimate + 1.96 * se,
    lower95 = estimate - 1.96 * se
  )

@@

To investigate the socio-demographics that had a strong effect on the two-party preferred vote, partial residual plots are used and shown in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Partial residuals, for a given variable, are the residuals from the fitted model with the estimated effect of that variable added to it. These plots show the direction, size and significance of an estimated effect, as well as any deviations from linearity. In each plot, the slope of the prediction line matches the estimated coefficient and the shaded region represents a 95% confidence band. Plots were computed using the method in @visreg. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. The estimated intercept is also added to the partial residuals for interpretability. Plots for each election are faceted to compare the effects over time in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Only socio-demographics that had a significant effect in at least two elections are displayed in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2).

@@

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as `Extractive` industries) were consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years (see row 1 in Figure \@ref(fig:partresplot1)). This is unsurprising, as the Liberal party had maintained close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (`Transformative`) were also more likely to support the Liberal party (see row 3 in Figure \@ref(fig:partresplot1)), from 2004 onwards.

@@

Of these job related variables, the most surprising effect is that associated with the proportion of workers in education, healthcare, social work, community and arts (`SocialServ`). Typically the Labor party has more generous funding schemes affecting these areas of work, so one might expect `SocialServ` to have had a negative effect on two-party preference. However, in every election this effect was found to be positive and significant.

@@

Typically the Labor party campaigns on more progressive policies, often including tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates were more likely to support the Liberal party, as the `Incomes` factor had a positive effect on Liberal preference (see row 6 in Figure \@ref(fig:partresplot1)). This effect was significant in every election aside from 2004 and 2010. `Unemployment` however, was not as influential. In 2001 and 2004, electorates with higher unemployment aligned with Labor, but over time this shifted towards support for the Liberal party, culminating in a positive (insignificant) effect in 2016.

@@

The older Australian population is often believed to be more conservative, and the left leaning political parties (including Labor) typically have a stronger appeal to younger people. This effect is indeed observed across the six elections, where populations between 20 and 34 years of age (`Pop_20_34`) aligned strongly with Labor preference (bottom row in Figure \@ref(fig:partresplot2)). Larger populations of 35 to 54 year olds (`Pop_35_54`) were also associated with Labor, but the magnitude of this effect was far smaller. Populations under 20 years of age was only significant in 2007, where `Pop_00_19` increased Liberal support.

@@

Larger migrant populations from Asia, the Middle East, South-Eastern Europe, the United Kingdom and elsewhere, were either associated with Labor support, or had no effect. Of these areas, only South-Eastern European populations significantly affected party preference, with larger populations associating with Labor in 2013 and 2016 (row 2, Figure \@ref(fig:partresplot2)). Speaking other languages (aside from English) however, appears to have had a far stronger effect, as observed through the `OtherLanguage` variable. Electorates with more diverse speech were associated with higher support for the Liberal party from 2004 onwards, with this effect being significant in 2013 and 2016 (see row 7, Figure \@ref(fig:partresplot1)). Furthermore, none of the variables relating to religious beliefs aside from Christianity had a material effect in any election (this includes the Buddhist, Muslim, Jewish, non-religious and Indigenous Australian populations). The association between Christian populations (`Christianity`) and the Liberal party steadily increased over the years, becoming positive and significant in 2016.

@@

In 2001, 2004 and 2016, higher proportions of people that recently (in the past five years) moved house (`DiffAddress`) increased electoral support for the Liberal party (see row 5 in Figure \@ref(fig:partresplot1)). This is somewhat surprising as one might expect house ownership and rental prices to be linked to two-party preference, rather than household mobility (`PropertyOwned` and `RentLoan` were not significant in any election).

@@

It is often hypothesised that states have a systematic bias towards one of the two major parties. Boxplots of residuals grouped by state (Figure \@ref(fig:resstate)) show that the data reflects this to only a limited extent. Tasmania and the Australian Capital Territory appeared to have a bias towards Labor, whereas the South Australia and the Northern Territory tended towards voting Liberal. However, there were relatively few electorates in each of these states (five, two, eleven and two respectively), so this apparent result may be due to incumbent effects rather than an actual state-specific bias.

```{r resstate, fig.cap = "Boxplot of residuals by state with jittered points. States ordered by median residual. A state-specific bias present only in the smaller states appears to have not been captured by the model.", fig.asp=0.6, out.width="70%", fig.pos = 'h', out.extra = ''}
all_resids %>%
  ggplot(aes(x = fct_reorder(StateAb, Residuals), y = Residuals)) +
  geom_hline(aes(yintercept = 0), col = "grey50") +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, shape = 1) +
  guides(col = FALSE, shape = FALSE, DivisionNm = FALSE) +
  theme_bw() +
  labs(x = "State", y = "Residuals") +
  theme(text = element_text(size = 16))
```

@@

The incumbent party appeared to have a distinct advantage at the next election. The boxplots in Figure \@ref(fig:resincumbent) show that if either of the Labor or Liberal parties won the seat at the previous election, the electorate is likely to vote in their favour, over and above any socio-demographic effects---this effect has not been captured by the model.

```{r resincumbent, fig.cap = "Boxplot of residuals for incumbent and non-incumbent parties each year. An incumbent advantage is evident and has not been captured by the model.", fig.asp=0.6, out.width="70%", fig.pos = 'h', out.extra = ''}
# Load first preferences
fp16 <- eechidna::fp16 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp13 <- eechidna::fp13 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp10 <- eechidna::fp10 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp07 <- eechidna::fp07 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp04 <- eechidna::fp04 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp01 <- eechidna::fp01 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))

@@

The electorate of Sydney (NSW) had a large Cook's distance and high leverage from 2001 to 2007, due to its diverse population (languages, birthplace and religion), high density of young adults (20 to 34 years old), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It remained a strong supporter of the Labor party and the extent of this support is underpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth was found to be an outlier in the 2013 and 2016 elections. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup---implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. In the elections since, the model underpredicts Wentworth's Liberal support.

@@

This paper explored the effects of electoral socio-demographic characteristics on the two-party preferred vote in the 2001--2016 elections, using information from the corresponding Australian federal elections and censuses. As a census did not always occur in the same year as an election, census data for each of the 2004--2013 elections were generated by employing a method of spatio-temporal imputation. This method imputes electoral socio-demographics for the electoral boundaries in place at the time of the corresponding election---an approach that is distinctly different from previous work on modelling election outcomes, where census and election data is typically joined without addressing their temporal differences. Before estimating a model, these socio-demographic variables were standardised (to adjust for changing variable scales) and subsets of variables (representing similar information) were combined into factors, resulting in a reduced predictor set. A spatial error model was then estimated for each election, accounting for the inherent spatial structure of the data.

@@

All election and census datasets, along with electoral maps and more, are available in the \textsf{eechidna} (Exploring Election and Census Highly Informative Data Nationally for Australia) \textsf{R} package, which can be downloaded from CRAN. The \textsf{eechidna} package makes it easy to look at the data from the Australian Federal elections and censuses that occurred between 2001 and 2016. This study contributed a large revision to the \textsf{eechidna} package, which included the addition of election and census data for 2001--2010, voting outcomes for polling booths and imputed census data for election years. For more details on using \textsf{eechidna}, please see the articles (vignettes) on the github page \url{ropenscilabs.github.io/eechidna}.

>>>>

The spatial autoregressive coefficient $\rho$ was positive and significant in the 2001, 2004 and 2016 elections (Figure \@ref(fig:rhovis)). In these three elections, there is evidence to suggest that neighbours shared some influential characteristics outside the explanatory variables, which affected the two-party preferred vote. Conversely, in the other three elections, the spatial effect was weaker and insignificant (although still positive).

```{r rhovis, fig.cap="Estimates of the spatial autoregressive parameter for each of the six elections, reported with their individual 95\\% confidence intervals. In 2001, 2004 and 2016 there was a significant spatial component.", out.height='25%', out.extra = ''}
rho_df <- bind_rows(
  glsmod16$rho_df %>% mutate(year = "2016"),
  glsmod13$rho_df %>% mutate(year = "2013"),
  glsmod10$rho_df %>% mutate(year = "2010"),
  glsmod07$rho_df %>% mutate(year = "2007"),
  glsmod04$rho_df %>% mutate(year = "2004"),
  glsmod01$rho_df %>% mutate(year = "2001")
) %>%
  mutate(
    upper95 = estimate + 1.96 * se,
    lower95 = estimate - 1.96 * se
  )

@@

To investigate the socio-demographics that had a strong effect on the two-party preferred vote, partial residual plots were used and shown in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). Partial residuals, for a given variable, are the residuals from the fitted model with the estimated effect of that variable added to it. These plots show the direction, size and significance of an estimated effect, as well as any deviations from linearity. In each plot, the slope of the prediction line matches the estimated coefficient and the shaded region represents a 95% confidence band. Plots were computed using the method in @visreg. If a horizontal line can be drawn through the confidence band, then the effect was insignificant. The estimated intercept was also added to the partial residuals for interpretability. Plots for each election are faceted in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2) to compare the effects over time. Only socio-demographics that had a significant effect in at least two elections are displayed.

@@

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as `Extractive` industries) were consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years (see row 1 in Figure \@ref(fig:partresplot1)). This is unsurprising, as the Liberal party maintained close ties with these traditional energy industries, and typically presented policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (`Transformative`) were also more likely to support the Liberal party (see row 3 in Figure \@ref(fig:partresplot1)), from 2004 onwards.

@@

Of these job related variables, the most surprising effect is that associated with the proportion of workers in education, healthcare, social work, community and arts (`SocialServ`). Typically the Labor party has had more generous funding schemes affecting these areas of work, so one might expect `SocialServ` to have had a negative effect on two-party preference. However, in every election this effect was found to be positive and significant.

@@

Typically the Labor party has campaigned on more progressive policies, often including tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates were more likely to support the Liberal party, as the `Incomes` factor had a positive effect on Liberal preference (see row 6 in Figure \@ref(fig:partresplot1)). This effect was significant in every election aside from 2004 and 2010. `Unemployment` however, was not as influential. In 2001 and 2004, electorates with higher unemployment aligned with Labor, but over time this shifted towards support for the Liberal party, culminating in a positive (insignificant) effect in 2016.

@@

The older Australian population has often been considered to be more conservative, and the left leaning political parties (including Labor) have typically had a stronger appeal to younger people. This effect was indeed observed across the six elections, where populations between 20 and 34 years of age (`Pop_20_34`) aligned strongly with Labor preference (bottom row in Figure \@ref(fig:partresplot2)). Larger populations of 35 to 54 year olds (`Pop_35_54`) were also associated with Labor, but the magnitude of this effect was far smaller. Populations under 20 years of age was only significant in 2007, where `Pop_00_19` increased Liberal support.

@@

Larger migrant populations from Asia, the Middle East, South-Eastern Europe, the United Kingdom and elsewhere, were either associated with Labor support, or had no effect. Of these areas, only South-Eastern European populations significantly affected party preference, with larger populations associating with Labor in 2013 and 2016 (row 2, Figure \@ref(fig:partresplot2)). Speaking other languages (aside from English) however, appears to have had a far stronger effect, as observed through the `OtherLanguage` variable. Electorates with more diverse speech were linked with higher support for the Liberal party from 2004 onwards, with this effect being significant in 2013 and 2016 (see row 7, Figure \@ref(fig:partresplot1)). Furthermore, none of the variables relating to religious beliefs aside from Christianity had a material effect in any election (this includes the Buddhist, Muslim, Jewish, non-religious and Indigenous Australian populations). The relationship between Christian populations (`Christianity`) and the Liberal party strengthened over the years, becoming positive and significant in 2016.

@@

In 2001, 2004 and 2016, higher proportions of people that recently (in the past five years) moved house (`DiffAddress`) increased electoral support for the Liberal party (see row 5 in Figure \@ref(fig:partresplot1)). This was somewhat surprising as one might expect house ownership and rental prices to be linked to two-party preference, rather than household mobility (`PropertyOwned` and `RentLoan` were not significant in any election).

@@

It is often hypothesised that states have a systematic bias towards one of the two major parties. Boxplots of residuals grouped by state (Figure \@ref(fig:resstate)) showed that the data reflects this to only a limited extent. Tasmania and the Australian Capital Territory appeared to have a bias towards Labor, whereas the South Australia and the Northern Territory tended towards voting Liberal. However, there were relatively few electorates in each of these states (five, two, eleven and two respectively), so this apparent result may be due to incumbent effects rather than an actual state-specific bias.

```{r resstate, fig.cap = "Boxplot of residuals by state with jittered points. States ordered by median residual. A state-specific bias present only in the smaller states appeared to have not been captured by the model.", fig.asp=0.6, out.width="70%", fig.pos = 'h', out.extra = ''}
all_resids %>%
  ggplot(aes(x = fct_reorder(StateAb, Residuals), y = Residuals)) +
  geom_hline(aes(yintercept = 0), col = "grey50") +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, shape = 1) +
  guides(col = FALSE, shape = FALSE, DivisionNm = FALSE) +
  theme_bw() +
  labs(x = "State", y = "Residuals") +
  theme(text = element_text(size = 16))
```

@@

The incumbent party appeared to have a distinct advantage at the next election. The boxplots in Figure \@ref(fig:resincumbent) show that if either of the Labor or Liberal parties won the seat at the previous election, the electorate was likely to vote in their favour at the subsequent election, over and above any socio-demographic effects---this effect has not been captured by the model.

```{r resincumbent, fig.cap = "Boxplot of residuals for incumbent and non-incumbent parties each year. An incumbent advantage was evident and had not been captured by the model.", fig.asp=0.6, out.width="70%", fig.pos = 'h', out.extra = ''}
# Load first preferences
fp16 <- eechidna::fp16 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp13 <- eechidna::fp13 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp10 <- eechidna::fp10 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp07 <- eechidna::fp07 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp04 <- eechidna::fp04 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))
fp01 <- eechidna::fp01 %>% 
  mutate(PartyAb = ifelse(PartyAb %in% c("LP", "NP"), "LNP", PartyAb))

@@

The electorate of Sydney (NSW) had a large Cook's distance and high leverage from 2001 to 2007, due to its diverse population (languages, birthplace and religion), high density of young adults (20 to 34 years old), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It remained a strong supporter of the Labor party and the extent of this support was underpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth was found to be an outlier in the 2013 and 2016 elections. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup---implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. In the elections that followed, the model underpredicted Wentworth's Liberal support.

@@

This paper explored the effects of electoral socio-demographic characteristics on the two-party preferred vote in the 2001--2016 elections, using information from the corresponding Australian federal elections and censuses. As a census did not always occur in the same year as an election, census data for each of the 2004--2013 elections were generated by employing a method of spatio-temporal imputation. This method imputes electoral socio-demographics for the electoral boundaries in place at the time of the corresponding election---an approach that is distinctly different from previous work on modelling election outcomes, where census and election data has typically been joined without addressing their temporal differences. Before estimating a model, these socio-demographic variables were standardised (to adjust for changing variable scales) and subsets of variables (representing similar information) were combined into factors, resulting in a reduced predictor set. A spatial error model was then estimated for each election, accounting for the inherent spatial structure of the data.

@@

All election and census datasets, along with electoral maps and more, are available in the \textsf{eechidna} (Exploring Election and Census Highly Informative Data Nationally for Australia) \textsf{R} package, which can be downloaded from CRAN. The \textsf{eechidna} package makes it easy to look at the data from the Australian Federal elections and censuses that occurred between 2001 and 2016. This study contributed a large revision to the \textsf{eechidna} package, which included the addition of election and census data for 2001--2010, voting outcomes for polling booths and imputed census data for election years. For more details on using \textsf{eechidna}, please see the articles (vignettes) on the github page: \url{ropenscilabs.github.io/eechidna}.
